Fish do not need Visas: the transboundary nature of the world’s exploited marine fish species (79/96 characters)

### 2020-01-07 Manuscript_Draft.Rmd
Updated manuscript draft and shared with co-authors

- **Potential Reviewers (J)**  
 -- Gakushi Ishimura,  
 -- Kathleen Miller, National Center for Atmospheric Research, Boulder, Colorado, USA  
 -- Malin Pinsky, Rutgers University, New Brunswick, New Jersey, USA  
 -- Megan Baily, 
 -- Steven D. Gaines, University of California, Santa Barbara, Santa Barbara, California, USA

- **Excluded Reviewers (J)**  
-- U. Rashid Sumaila, Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada  
-- Louise S.L. Teh, Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada

@@

@@

## Data and materials availability
*Any restrictions on materials such as MTAs.*

Gabriel, is there any restriction with the data?

<!-- # Materials and Methods -->

<!-- We determine whether or not a species can be considered transboundary. The method is based on a series of criteria to be met and the overlay of species distributions and the spatial boundaries of the world’s EEZs. For the current analysis we work at the species level and adopt the Food and consider transboundary species as those that occur within the Exclusive Economic Zone (EEZ) of neighboring countries [@Munro:2002uf]. All data was scaled to 0.5 degree latitude x 0.5 degree longitude grid cells (*n* = ~180,000 grid cells). -->

<!-- ## Databases on species geographic distributions -->

<!-- To determine the number of transboundary marine species exploited by fisheries within each of the world's EEZs we first extracted all the commercial marine fish or invertebrate species from the *Sea Around Us* database (http://www.seaaroundus.org and determined their current distributions. For this, we used four data sources of species-distributions: (*i*) observational data, (*ii*) an Environmental Niche Model (ENMs), (*iii*) a life-history-based distribution model, and (*iv*) fisheries catch data (**Table x**). Each source represent a different method of estimating the distribution of a given species and thus, provides a more robust result. Only commercial fished species with data from all four sources were included in the analysis. The final dataset comprised **XXX** species, **XX**% of the reported marine species in the SAU database and **XX**% of the reported taxa of the FAO (**See S1_Data**). -->

<!-- ### *Occurence data* -->

<!-- The occurrence data was collected from five publicly available repositories; Fishbase (http://fishbase.org), the Global Biodiversity Information Facility (GBIF; https://www.gbif.org/), the Ocean Biogeographic Information System (OBIS; https://obis.org/),  the Intergovernmental Oceanographic Comission (IOC; http://ioc-unesco.org), and the International Union for Conservation of Nature (IUCN; https://www.iucn.org/technical-documents/spatial-data) [@Reygondeau:2019uh].   -->

<!-- ### *Distribution models* -->

<!-- In addition to the occurrence data we use two different methods to estimate species distributions, hereafter referred as ENM-Nereus and SDM-SAU. Although they use the same data, the models are structurally different complementing each other and providing robustness to the results. -->

<!-- The ENM-Nereus consists in a multimodel approach based on a Bioclim and a Bososted Regression Tree model [@Thuiller:2009gp], a Maxent model [@Phillips:2006ff], and a Non-Parametric Probabilistic Ecological Niche Model [@Beaugrand:2011fd]. Environmental variables utilized in the models included sea surface temperature, surface pH, surface oxygen concentration, and vertically integrated (0–100 m) net primary production (NPP) [@Asch:2018ca]. The ENM-Nereus employed all of the observational data-sources previously mentioned. Global environmental conditions were obtained from 1970 to 2000 and average from three Earth System Models developed by the Geophysical Fluid Dynamics Laboratory (GFDL- https://www.gfdl.noaa.gov/earth-system-model/), the Institute Pierre Simon Laplace (IPSL- www.icmc.ipsl.fr/), and the Max Planck Institute for Meteorology (MPI- www.mpimet.mpg.de/en/science/models/). See [@Asch:2018ca; @Reygondeau:2019uh] for model details. -->

<!-- The SDM-SAU model follows a five-steps process based on species-specific life history information, rather than environmental variables [@Close:2006ux; @Palomares:wg]. For each commercial fish species, the model first uses the FAO major fishing areas and countries EEZs to determine a broad distribution. It then uses life history information to delimit its range within the FAO fishing area (e.g. thermal preference, depth limits). The range is delimited even further by expert-review polygons and compared with that of AquaMaps [@Kaschner:2016tl], OBIS and GBIF occurrence data. The model then determines a species habitat preference based on the assumptions that the relative abundance of a species is determined by the number of habitats in a grid cell and the distance of the species to each habitat, as well as the importance of the habitat to the species size. Finally, the species equatorial submergence (e.g. the latitudinal region where a species is not seen in between poles) is estimated for each species. See [@Close:2006ux; @Palomares:wg] for model details. -->

<!-- ### *Catch data* -->

<!-- The previous models combine observational data with a series of biotic and abiotic information to determine the probability that a species will be found in a given space at a given time. However, this does not mean that the species in question will actually be there. While the models do use approaches to double-check species occurrences (e.g. ENM-Nereus uses four different species distribution algorithms and SPD-SAU undertakes validation by means of other models), we used a fourth data source to corroborate the models’ outputs. The *Sea Around Us* estimates total reconstructed catches - i.e., catches  based on all publicly available information sources and including discards, as well as unreported and illegal catches that are not included in the FAO data available for each country. Catches are also spatially allocated [@Zeller2016]. Thus, we used the Sea Around Us catch reconstruction database from 2005 to 2014 (last ten years of data) as the fourth dataset to estimate transboundary species and to estimate their catch contribution within EEZs. -->

@@

<!-- ## Determine if a species is transboundary -->

<!-- We developed a four-criteria method for determining whether or not a species was transboundary. Only species that meet all criteria were considered as transboundary species and the analysis was done only within the EEZs of coastal states. In some cases, criteria also work as indexes to measure the uncertainty in the analysis as they provided continuous index (e.g. from 0 to 1) rather than a discrete, yes or no, index . All of the analysis was done in the statistical software *R version 3.5.2 (2018-12-20)* with the packages *data.table*[@Packagedatatable:2019uh], *janitor*[@PackagejanitorSim:2018ur], *wesanderson*[@Packagewesanderson:2018vn], *rfishbase*[@PackagerfishbaseR:2019th], *R.matlab*[@PackageRmatlabRe:2018un], *sf*[@PackagesfSimpleF:2018vp], *sp*[@PackagespClasses:2019tn], *tidiverse*[@PackagetidyverseE:2017vq], *tidytext*[@PackagetidytextTe:2019vk], and *zoo*[@PackagezooSInfr:2019tm]. All code is available at https://github.com/jepa/FishForVisa. -->

<!-- ### Criteria 1; Neighboring EEZs -->

<!-- As previously mentioned, we define transboundary species as those marine species that happen within the EEZs of two or more neighboring countries. Hence, the first criteria was that the species would only be transboundary between two neighboring countries, regardless of the species extended distribution. This way, a species such as Atlantic cod (*Gadus morhua*) distributed along the north Atlantic, was only considered transboundary between each of the neighboring nations covering its distribution, rather than the region as a whole. We assume that this would reduce the differences between species and stock as is more likely that neighboring species belong to the same stock, rather than, lets say, Cod in Norway and Canada. However, the analysis here presented was kept at the species level and stocks within countries were not considered. -->

<!-- We define the EEZ boundaries using the SAU shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) that sub-divides EEZs by regions (e.g. Mexico Pacific and Mexico Atlantic) and determine the intersections between polygons using the *sf*-*R* package. When estimating transboundary species, we filter out those shared by the EEZs sub-regions, and when aggregating by country we only counted the species once, if it appeared in more than one sub-region. Species that were present in EEZs that were non-continental territories neighboring other countries were kept (e.g. Argentina and Falkland Islands), but removed when the non-continental territory belonged to the same nation (e.g. Brazil and Fernando de Noronha). -->

<!-- ### Criteria 2; Data agreement -->

<!-- We use the occurrence database, the ENM-Nereus model, and SDM-SAU model to determine the presence of each species within each of the 0.5 x 0.5 grid cells. We computed a *Species Index* by dividing the number of datasets that confirm presence of a species in a grid cell over the total amount of datasets (*n* = 3). The index is unit-less and represents a scale from 0.3 to 1 where 0.3 means only one dataset reports presence of the species in a grid cell, 0.6 represents two of three sources, and 1 means agreement between all datasets. For the analysis, we selected only those cases where *Species Index* = 1 to get a more conservative estimate of transboundary species.  -->

<!-- ### Criteria 3; Modeling verification -->

<!-- We assume that a species was only present in a given grid-cell if it was reported in the SAU catch database. Therefore, all species that were not reported as caught in any single year between the reference years (2005 to 2014) in a given grid-cell were dropped, regardless of the *Species Index*. The assumption relies in that if a commercial species is projected within any fishing country, such species would have been fished, and thus reported at some period of time, thus validating the models. -->

<!-- ### Criteria 4; Spatial distribution -->

<!-- Finally, in order to have a more robust result and do not determine a transboundary species based on the presence in a single 0.5 x 0.5 grid cell within an EEZ, we computed an *Area Index*. The *Area Index* consist in the proportion of the total species distribution within both EEZs that each neighboring EEZ has. We determine that a species would be transboundary if both neighboring EEZs enclosed over 25% of the species joint distribution. Such threshold can be lowered for a more relaxed result or increased for a more conservative estimate (**S2_Hist_per_Treshold**). -->

<!-- ## Fisheries trends -->

<!-- We estimated the economic contribution of transboundary species for each country using global ex-vessel price [@Tai2017]. The data includes ex-vessel price from multiple sources and a structured interpolation method (e.g. similar countries, species) to fill in data gaps [@Sumaila:2015uc]. The contribution of transboundary species was done by aggregating the catch of each species that happened within the EEZ of the countries sharing that species. For this study, we did not include catch nor revenue from areas beyond national jurisdiction. Finally, we use the catch data to determine the exploitation category of each species within a EEZ. Although this method has previously used to estimate stock status [@Grainger:1996tk], the categories presented here are intended to be seen as catch trends, and not status of each species. We only assessed species within each EEZ that had at least ten years of data between the first and last reported landings with at least five consecutive data years. Three categories were drawn from a method depending on the year's catch, the maximum catch over time and the minimum catch after the catch peak (**Table XX**) [@Kleisner:2011wn]. Finally, we reported the predominant category over the last ten years of catch data (2005-2014). We perform a series of one-way analysis of variance ( one-way ANOVA) to determine significant differences (statistically) between the means of different groups (e.g. categories, regions, species) regarding transboundary species and their contribution in terms of catch and revenue. -->

@@

## Caveats

@@

@@

# References (59/about 30)

>>>>

- **Potential Reviewers (J)**  
 -- Gakushi Ishimura, Faculty of Agriculture Department of Food Production and Environmental Management, Iwate University, Morioka, Japan  
 -- Kathleen Miller, National Center for Atmospheric Research, Boulder, Colorado, USA  
 -- Louise S.L. Teh, Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada  
 -- Malin Pinsky, Rutgers University, New Brunswick, New Jersey, USA  
 -- Megan Baily, Integrated Ocean and Coastal Governance, Dalhousie University, Halifax, Canada  
 -- Steven D. Gaines, University of California, Santa Barbara, Santa Barbara, California, USA

- **Excluded Reviewers (J)**

@@

Is worth mention that for the current study we did not consider climate change as a driver of species' distribution. Dsitribution shifts are the most recorded response of marine species to the effects of climate change in the ocean [@Poloczanska:2016kk]. Moreover, catch composition might have already changed in the last five decades [@Cheung:2013gk], and will most likeley continue throuugh the 21^st century [@Cheung:2016jd]. While considering climate change in the distribution of shared stocks is beyond the scope of this study, it is likely that it will bring at least one new transboundary species per EEZ in the future [@Pinsky:2018cb], while also change the proportion of the distribution (e.g. Area Index) that each county's share to a degree that could influecne fisheries management [@Sumailaetal_in_press]. Exploring the impacts of climate change in share species is key to develop adaptation methods to minimize the effect of such distributional change on the conservation of species as well as the economy and food security of coastal nations.

@@

## Data and materials availability

All data is available by request of the authors.

# Materials and Methods

We determine whether or not a species can be considered transboundary. The method is based on a series of criteria to be met and the overlay of species distributions and the spatial boundaries of the world’s EEZs. For the current analysis we work at the species level and consider transboundary species as those that occur within the Exclusive Economic Zone (EEZ) of neighboring countries [@Munro:2002uf]. All data was scaled to $0.5^o$ x $0.5^o$ lat-long grid cells (*n* = ~180,000 grid cells).

## Databases on species geographic distributions

To determine the number of transboundary marine species exploited by fisheries within each of the world's EEZs we used the list of commercial marine fish and invertebrate species from the *Sea Around Us* (SAU) database (http://www.seaaroundus.org) and determined their current distributions. For this, we used four data sources of species-distributions: (*i*) observational data, (*ii*) an Environmental Niche Model (ENMs), (*iii*) a life-history-based distribution model, and (*iv*) fisheries catch data (**Table x**). Each source represent a different method of estimating the distribution of a given species and thus, provides a more robust result. Only commercial fished species with data from all four sources were included in the analysis. The final dataset comprised 968 marine species (including andornomus species like salmons), **XX**% of the reported marine species in the SAU database and **XX**% of the reported taxa of the FAO (**See S1_Species_List**).

### *Occurence data*

The occurrence data was collected from five publicly available repositories; Fishbase (http://fishbase.org), the Global Biodiversity Information Facility (GBIF; https://www.gbif.org/), the Ocean Biogeographic Information System (OBIS; https://obis.org/),  the Intergovernmental Oceanographic Comission (IOC; http://ioc-unesco.org), and the International Union for Conservation of Nature (IUCN; https://www.iucn.org/technical-documents/spatial-data) [@Reygondeau:2019uh].

### *Distribution models*

In addition to the occurrence data we use two different methods to estimate species distributions, hereafter referred as ENM-Nereus and SDM-SAU. Although they use the same data, the models are structurally different complementing each other and providing robustness to the results.

The ENM-Nereus consists in a multimodel approach based on a Bioclim and a Bososted Regression Tree model [@Thuiller:2009gp], a Maxent model [@Phillips:2006ff], and a Non-Parametric Probabilistic Ecological Niche Model [@Beaugrand:2011fd]. Environmental variables utilized in the models included sea surface temperature, surface pH, surface oxygen concentration, and vertically integrated (0–100 m) net primary production (NPP) [@Asch:2018ca]. The ENM-Nereus employed all of the observational data-sources previously mentioned. Global environmental conditions were obtained from 1970 to 2000 and average from three Earth System Models developed by the Geophysical Fluid Dynamics Laboratory (GFDL- https://www.gfdl.noaa.gov/earth-system-model/), the Institute Pierre Simon Laplace (IPSL- www.icmc.ipsl.fr/), and the Max Planck Institute for Meteorology (MPI- www.mpimet.mpg.de/en/science/models/). See [@Asch:2018ca; @Reygondeau:2019uh] for model details.

The SDM-SAU model follows a five-steps process based on species-specific life history information, rather than environmental variables [@Close:2006ux; @Palomares:wg]. For each commercial fish species, the model first uses the FAO major fishing areas and countries EEZs to determine a broad distribution. It then uses life history information to delimit its range within the FAO fishing area (e.g. thermal preference, depth limits). The range is delimited even further by expert-review polygons and compared with that of AquaMaps [@Kaschner:2016tl], OBIS and GBIF occurrence data. The model then determines a species habitat preference based on the assumptions that the relative abundance of a species is determined by the number of habitats in a grid cell and the distance of the species to each habitat, as well as the importance of the habitat to the species size. Finally, the species equatorial submergence (e.g. the latitudinal region where a species is not seen in between poles) is estimated for each species. See [@Close:2006ux; @Palomares:wg] for model details.

### *Catch data*

The previous models combine observational data with a series of biotic and abiotic information to determine the probability that a species will be found in a given space at a given time. However, this does not mean that the species in question will actually be there. While the models do use approaches to double-check species occurrences (e.g. ENM-Nereus uses four different species distribution algorithms and SPD-SAU undertakes validation by means of other models), we used a fourth data source to corroborate the models’ outputs. The *Sea Around Us* estimates total reconstructed catches - i.e., catches  based on all publicly available information sources and including discards, as well as unreported and illegal catches that are not included in the FAO data available for each country. Catches are also spatially allocated [@Zeller2016]. Thus, we used the Sea Around Us catch reconstruction database from 2005 to 2014 (last ten years of data) as the fourth dataset to estimate transboundary species and to estimate their catch contribution within EEZs.

@@

## Determine if a species is transboundary

We developed a four-criteria method for determining whether or not a species was transboundary. Only species that meet all criteria were considered as transboundary species and the analysis was done only within the EEZs of coastal states. In some cases, criteria also work as indexes to measure the uncertainty in the analysis as they provided continuous index (e.g. from 0 to 1) rather than a discrete, yes or no, index . All of the analysis was done in the statistical software *R version 3.5.2 (2018-12-20)* with the packages *data.table*[@Packagedatatable:2019uh], *janitor*[@PackagejanitorSim:2018ur], *wesanderson*[@Packagewesanderson:2018vn], *rfishbase*[@PackagerfishbaseR:2019th], *R.matlab*[@PackageRmatlabRe:2018un], *sf*[@PackagesfSimpleF:2018vp], *sp*[@PackagespClasses:2019tn], *tidiverse*[@PackagetidyverseE:2017vq], *tidytext*[@PackagetidytextTe:2019vk], and *zoo*[@PackagezooSInfr:2019tm]. All code is available at https://github.com/jepa/FishForVisa.

### Criteria 1; Neighboring EEZs

As previously mentioned, we define transboundary species as those marine species that happen within the EEZs of two or more neighboring countries. Hence, the first criteria was that the species would only be transboundary between two neighboring countries, regardless of the species extended distribution. We define the EEZ boundaries using the SAU shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) that sub-divides EEZs by regions (e.g. Mexico Pacific and Mexico Atlantic) and determine the intersections between polygons using the *R* package *sf*. When estimating transboundary species, we filter out those shared by the EEZs sub-regions, and when aggregating by country we only counted the species once, if it appeared in more than one sub-region. Species that were present in EEZs that were non-continental territories neighboring other countries were kept (e.g. Argentina and Falkland Islands), but removed when the non-continental territory belonged to the same nation (e.g. Brazil and Fernando de Noronha). **Sub-states** like Puerto Rico and Hong Kong were added to their **parent** state.

### Criteria 2; Data agreement

We use the occurrence database, the ENM-Nereus model, and SDM-SAU model to determine the presence of each species within each of the 0.5 x 0.5 grid cells. We computed a *Species Index* by dividing the number of datasets that confirm presence of a species in a grid cell over the total amount of datasets (*n* = 3). The index is unit-less and represents a scale from 0.3 to 1 where 0.3 means only one dataset reports presence of the species in a grid cell, 0.6 represents two of three sources, and 1 means agreement between all datasets. For the analysis, we selected only those cases where *Species Index* = 1 to get a more conservative estimate of transboundary species.

### Criteria 3; Modeling verification

We assume that a species was only present in a given grid-cell if it was reported in the SAU catch database. Therefore, all species that were not reported as caught in any single year between the reference years (2005 to 2014) in a given grid-cell were dropped, regardless of the *Species Index*. The assumption relies in that if a commercial species is projected within any fishing country, such species would have been fished, and thus reported at some period of time, thus validating the models.

### Criteria 4; Spatial distribution

Finally, in order to have a more robust result and do not determine a transboundary species based on the presence in a single 0.5 x 0.5 grid cell within an EEZ, we computed an *Area Index*. The *Area Index* consist in the proportion of the total species distribution within both EEZs that each neighboring EEZ has. We determine that a species would be transboundary if both neighboring EEZs enclosed over 25% of the species joint distribution. Such threshold can be lowered for a more relaxed result or increased for a more conservative estimate (**S2_Hist_per_Treshold**).

## Fisheries trends

We estimated the economic contribution of transboundary species for each country using global ex-vessel price [@Tai2017]. The data includes ex-vessel price from multiple sources and a structured interpolation method (e.g. similar countries, species) to fill in data gaps [@Sumaila:2015uc]. The contribution of transboundary species was done by aggregating the catch of each species that happened within the EEZ of the countries sharing that species. For this study, we did not include catch nor revenue from areas beyond national jurisdiction. Finally, we use the catch data to determine the exploitation category of each species within a EEZ. Although this method has previously used to estimate stock status [@Grainger:1996tk], the categories presented here are intended to be seen as catch trends, and not status of each species. We only assessed species within each EEZ that had at least ten years of data between the first and last reported landings with at least five consecutive data years. Three categories were drawn from a method depending on the year's catch, the maximum catch over time and the minimum catch after the catch peak (**Table XX**) [@Kleisner:2011wn]. Finally, we reported the predominant category over the last ten years of catch data (2005-2014). We perform a series of one-way analysis of variance ( one-way ANOVA) to determine significant differences (statistically) between the means of different groups (e.g. categories, regions, species) regarding transboundary species and their contribution in terms of catch and revenue.

@@

## Analysis Assumptions and Caveats
<!-- # Related to Criteria 1 (Neighbours) -->
We ran the analysis at the species level due to limited spatial-specific data on species sub populations (stocks), therefore we are not able to identify transboundary stocks within EEZs. However, we limited the distribution to neighbouring countries. This way, a species such as Atlantic cod (*Gadus morhua*) distributed along the north Atlantic, was only considered transboundary between each of the neighboring nations, rather than the distribution as a whole. We assume that this would reduce the gap between species and stock, as is more likely that neighboring species belong to the same stock, rather than, lets say, Cod in Norway and Canada.
<!-- Related to Criteria 2 (Models) -->
The models implemented in this analisis are not free of uncertainty. Structural differences within Earth System Models results in variations of oceanic conditions, which undoubtedly permeate to the ENM-Nereus. We averaged the results from the three models in an effort to capture the structural variation of the models. Climate natural variability is a mayor driver of marine species distribution, potentially removing a species from an EEZ for long periods (e.g. Anchovies and Sardines "regimes" in the Eastern Pacific are strongly influenced by water temperature decal oscillations [@Chavez:2003jr]). Thus, a species' distribution can potentially be reduced or shifted in such way that it only covers one EEZ until the proper oceanic conditions re-expands its distribution. To account for such climate variability, we obtained the ESM-Nereus results as an average of oceanic conditions between 1970 and 2000. This is not an issue for the SDM-SAU as it does not directly require environmental variables [@Palomares:wg]. Moreover, the occurrence data is non-time dependant, meaning that we collected all observational data available, regarding of the date when the species was recorded.
<!-- Related to Criteria 3 (SAU catch data) -->
In addition, we assumed that, if the SAU reconstruction data recorded a species caught in any given grid cell, then the species was present in that grid cell. However, catch data is not exempt of uncertainty [@]. Thus, our catch data reference period in any given grid cell was between 2005 and 2014, the last ten years of data.
<!-- Related to Criteria 4; Spatial distribution -->
Finally, the results of this study only considered those species that had a *Species Index* of 1 (e.g. all datasets agree on presence) and at least 25% of *Area Index* (e.g. the species shared distribution was at least 25% in each EEZ). This presents a relative conservative estimate of the number of transboundary species in the world. However, using a different *Species Index* will give results based on less datasets while a smaller *Area Index* will estimate more transboundary species (S3_Area.png).

@@

img <- readPNG(here("./Figures/S3_Area.png"))
grid.raster(img)
```

@@

# References (37/about 30)
